---
title: "How scene ordering works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How scene ordering works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # These chunks talk to the live M2M API, which needs credentials, so
  # nothing is evaluated at build time.
  eval = FALSE
)
```

The main vignette shows the pipeline as one chain. This one explains what is
actually happening underneath: why getting a scene takes several steps, where
scene lists fit, and what the download queue is doing while you wait.

```{r setup}
library(USGSm2m)
```

## Why it takes more than one call

There is no single "download this scene" endpoint. Four separate things have to
be established before the API will hand over a file:

1. **Which scenes** you mean — found by searching a dataset.
2. **A server-side name for that set** — a *scene list*, because the endpoints
   in step 3 will not accept scene ids inline.
3. **Which products** of those scenes you want — a scene is not one file but a
   bundle, its individual bands, and browse images.
4. **A prepared download** — the distribution system may have to stage a
   product before there is a URL to fetch.

Each step has its own endpoint, and each depends on the one before:

```
   scene-search          which scenes exist
        │
        ▼
   scene-list-add        name that set server-side  ──►  listId
        │
        ▼
   download-options      what is downloadable       (requires listId)
        │
        ▼
   download-request      queue it                   ──►  label
        │
        ▼
   download-retrieve     collect URLs               (by label)
        │
        ▼
   HTTP GET              the actual bytes
```

In this package steps 1–3 are `$search()`, `$scene_list()` and `$products()`,
and steps 4–5 are `$request()` and `$retrieve()`. `$products()` performs step 2
for you, which is why the scene list is usually invisible.

## Scene lists

### What registering actually does

A scene list is not something the package holds in memory. It is a **named set
of scene ids stored on the USGS server**, created by posting to
`scene-list-add`:

```
{
  "listId":      "USGSm2m_20260821150915_vbbem6",   <- a name you invent
  "idField":     "entityId",
  "entityIds":   ["LC80430322020199LGN00", ...],
  "datasetName": "landsat_ot_c2_l2"
}
```

You choose the identifier, the server stores the set under it, and the call
reports how many scenes the list now holds. From then on that id stands in for
the whole set. It is closer to uploading a temporary table than to creating an
object locally.

`download-options` and `scene-metadata-list` will not accept scene ids inline —
only a `listId`. For a search returning thousands of scenes that avoids
re-sending every id on each call, but it means even a single-scene download has
to go through the same ceremony.

### What the package does for you

Written out by hand, the registration step looks like this: invent a unique id,
add the scenes under it, then pass the same id *and the same dataset name* to
the next call.

```{r}
list_id <- "my_unique_id_12345"
ers_scene_list_add(session, "landsat_ot_c2_l2", scenes, list_id)
products <- ers_scene_products(session, "landsat_ot_c2_l2", list_id)
```

Three things have to stay consistent there, and each fails in its own way: a
reused identifier silently appends to an existing set, a mismatched
`datasetName` fails obscurely, and the id has to be threaded between the calls.

`$products()` collapses all of that into one line, because the search object
already knows both the scenes and the dataset:

```{r}
found$products()
```

Internally it is still the same two calls — `$products()` asks
`$scene_list()` to register, then requests download options against the
resulting id. The identifier is generated (a timestamp plus six random
characters) so it cannot collide, and the dataset name is carried on the object
rather than retyped, so it cannot disagree.

### One registration per search

Repeated calls reuse the same list rather than registering the scenes again:

```{r}
found$products()
found$products()
found$scene_list()$list_id   # all three refer to one registration
```

A narrowed search is a different set of scenes, so it gets its own:

```{r}
found$filter(cloudCover < 10)$scene_list()   # separate list
```

Passing an explicit `list_id` always registers under that name and leaves the
reused one alone, and a list you have deleted with `$remove()` is registered
afresh next time rather than being handed back dead.

Note that nothing deletes these lists automatically — they persist until they
expire, or until you call `$remove()`.

### The handle

`M2MSceneList` is not the scene list itself — it is a **reference to one living
on the server**. That is why `$remove()` deletes it remotely, and why the object
keeps working after the list expires: `$scenes()` simply returns zero rows.

The class exists for three reasons: it is what `$products()` produces
internally, so the step is representable rather than buried; it is the only way
to reach the endpoints that work in bulk; and it lets you reattach to a list
registered earlier.

`$scene_list()` registers the scenes from a search and returns that handle:

```{r}
found <- sess$dataset("landsat_ot_c2_l2")$search(
  temporal = filter_temporal("2020-07-01", "2020-07-31"),
  max_results = 4
)

scenes <- found$scene_list()
scenes
#> <M2MSceneList>
#>   List id: USGSm2m_20260821150915_vbbem6
#>   Dataset: landsat_ot_c2_l2
#>   Next:    $products()  |  $metadata()  |  $summary()
```

The identifier is generated for you. You can supply your own to
`$scene_list(list_id = )` if you want a stable, memorable name to come back to.

### How adding behaves

Three behaviours are worth knowing, because none is obvious from the endpoint
name:

* Adding **appends** to a list rather than replacing its contents.
* The call reports the list's **cumulative** size, not how many you just added.
* Adding a scene already in the list is a **no-op**, so retries are safe.

Adding two scenes, then two more, then re-adding the first two, gives a list of
four:

```
add first 2     -> reports 2   list holds 2
add next 2      -> reports 4   list holds 4
re-add first 2  -> reports 4   list holds 4
```

### Lifetime

Scene lists are temporary. They expire server-side after a period of
inactivity, and `$summary()` reports a `listTimeout` for lists that carry one.
Nothing breaks when a list expires — it simply stops being found, and
`$scenes()` returns zero rows rather than raising an error. The same is true
after you delete one with `$remove()`.

This matters when resuming work later: **a scene list will usually be gone by
the next session, but the download order it produced will not.** Reconnect to
orders by label, not to scene lists by id.

### Doing something with the list itself

Most of the time you go straight to `$products()`. The list is worth holding
when you want the endpoints that only work in bulk:

```{r}
scenes$metadata()   # every scene's full metadata in one call
scenes$summary()    # combined spatial and temporal extent of the set
scenes$scenes()     # the entity ids currently in the list
```

`$metadata()` is the main reason to care: fetching metadata scene-by-scene
would be one request each.

If you reattach to a list by id, the package does not initially know which
dataset it holds — the API only reveals that through the list's summary — so
`$dataset()` resolves it on demand:

```{r}
again <- sess$scene_list("USGSm2m_20260821150915_vbbem6")
again
#>   Dataset: <unresolved>
again$dataset()
#> [1] "landsat_ot_c2_l2"
```

Pass `dataset_name` to `$scene_list()` to skip that lookup, or to choose when a
list spans more than one dataset.

## Choosing products

A scene is not a file. `download-options` reports several *products* per
scene, at two different granularities:

* **Whole products** — a Level-1 Product Bundle (one `.tar` containing the
  entire scene), or a full-resolution browse image. Listed by
  `$scene_products()`.
* **Individual files** — the separate bands and metadata files nested inside a
  bundle. Listed by `$bands()`.

For one Landsat Collection 2 scene that is roughly ten whole products and
nineteen individual files. Browse images have no constituent files, so they
appear only in `$scene_products()`.

Pick a granularity with the matching selector, then narrow with `$filter()`:

```{r}
opts <- found$products()

opts$select_bands(c("B4", "B5"))$filter(bulkAvailable)   # 6 separate .TIFs
opts$select_products("Product Bundle")$filter(available) # 1 .tar per scene
```

`$selected()` always shows what `$request()` will queue. Neither selector
filters on availability, because the API lists superseded products with
`available = FALSE` and silently dropping them would hide that they exist.

## The download queue

`$request()` places an order under a label. What comes back depends on whether
the distribution system can serve the products immediately:

* Products already staged return **with URLs**, in `$available`.
* Anything needing preparation is accepted and reported in `$requested`, with
  no URL yet.

`$is_ready()` tells you which situation you are in. `$refresh()` re-polls and
adds newly ready files to `$available`:

```{r}
queue <- opts$select_bands(c("B4", "B5"))$request(label = "ndvi_july_2020")

while (!queue$is_ready()) {
  Sys.sleep(30)
  queue$refresh()
}

queue$retrieve("data/landsat")
```

`$refresh()` accumulates rather than replaces, because the API drops entries
once it has handed them over — a plain replacement would lose URLs collected on
an earlier poll.

Some orders are staged rather than queued, and need `$prepare()` to start
processing. Along with `$cancel()`, it is one of the two methods here that
change server-side state.

## Coming back later

Orders outlive both your R session and the scene list that produced them, so a
large download does not have to finish in one sitting:

```{r}
sess <- m2m_session()

sess$download_labels()      # which orders exist
queue <- sess$download_queue("ndvi_july_2020")
queue$refresh()
queue$retrieve("data/landsat")
```

`sess$downloads()` lists every individual download across all labels, and
`queue$summary()` breaks one order down by dataset.

## What lives where

| Thing | Identified by | Survives your session? | Created by |
|---|---|---|---|
| Session | API token | No — expires when idle | `m2m_session()` |
| Scene list | `listId` | Usually not — expires when idle | `$scene_list()`, or implicitly by `$products()` |
| Download order | `label` | Yes | `$request()` |
| Downloaded files | path on disk | Yes | `$retrieve()` |

The practical consequence: give orders labels you will recognise later, and do
not expect a scene list to still be there tomorrow.
