---
title: "How scene ordering works"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How scene ordering works}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # These chunks talk to the live M2M API, which needs credentials, so
  # nothing is evaluated at build time.
  eval = FALSE
)
```

The main vignette shows the pipeline as one chain. This one explains what is
actually happening underneath: why getting a scene takes several steps, where
scene lists fit, and what the download queue is doing while you wait.

```{r setup}
library(USGSm2m)
```

## Why it takes more than one call

There is no single "download this scene" endpoint. Four separate things have to
be established before the API will hand over a file:

1. **Which scenes** you mean — found by searching a dataset.
2. **A server-side name for that set** — a *scene list*, because the endpoints
   in step 3 will not accept scene ids inline.
3. **Which products** of those scenes you want — a scene is not one file but a
   bundle, its individual bands, and browse images.
4. **A prepared download** — the distribution system may have to stage a
   product before there is a URL to fetch.

Each step has its own endpoint, and each produces exactly the one thing the
next step needs:

```
  scene-search        which scenes match
       │
       │  entityIds
       ▼
  scene-list-add      register that set server-side
       │
       │  listId
       ▼
  download-options    what is downloadable for those scenes
       │
       │  entityId + productId pairs
       ▼
  download-request    place the order
       │
       │  label
       ▼
  download-retrieve   collect prepared URLs
       │
       │  urls
       ▼
  HTTP GET            the bytes
```

Follow the labels down the middle and the whole thing is a chain of handoffs:
a search gives you scene ids, registering them gives you a `listId`, that gives
you product ids, ordering those gives you a `label`, and the label eventually
gives you URLs.

Each endpoint maps onto one method:

| API endpoint | Package | Needs | Gives you |
|---|---|---|---|
| `scene-search` | `$search()` | filters | `M2MSceneSearch` — `$scenes`, `$total_hits` |
| `scene-list-add` | `$scene_list()` | scenes, dataset | `M2MSceneList` — holds the `listId` |
| `download-options` | `$products()` | `listId`, dataset | `M2MDownloadOptions` — `$bands()`, `$scene_products()` |
| `download-request` | `$request(label)` | entityId/productId pairs | `M2MDownloadQueue` — `$ready()`, `$pending()` |
| `download-retrieve` | `$refresh()` | `label` | updates `$available` in place |
| (plain HTTP GET) | `$retrieve(dir)` | URLs | tibble of `path` and `status` per file |

Two of these you will rarely call yourself. `$products()` performs
`scene-list-add` as well as `download-options`, which is why the scene list is
usually invisible, and `$request()` already returns whatever was available
immediately, so `$refresh()` only matters when something needed staging.

## Scene lists

### What registering actually does

A scene list is not something the package holds in memory. It is a **named set
of scene ids stored on the USGS server**, created by posting to
`scene-list-add`:

```
{
  "listId":      "USGSm2m_20260821150915_vbbem6",   <- a name you invent
  "idField":     "entityId",
  "entityIds":   ["LC80430322020199LGN00", ...],
  "datasetName": "landsat_ot_c2_l2"
}
```

You choose the identifier, the server stores the set under it, and the call
reports how many scenes the list now holds. From then on that id stands in for
the whole set. It is closer to uploading a temporary table than to creating an
object locally.

`download-options` and `scene-metadata-list` will not accept scene ids inline —
only a `listId`. For a search returning thousands of scenes that avoids
re-sending every id on each call, but it means even a single-scene download has
to go through the same ceremony.

### What the package does for you

Written out by hand, the registration step looks like this: invent a unique id,
add the scenes under it, then pass the same id *and the same dataset name* to
the next call.

```{r}
list_id <- "my_unique_id_12345"
ers_scene_list_add(session, "landsat_ot_c2_l2", scenes, list_id)
products <- ers_scene_products(session, "landsat_ot_c2_l2", list_id)
```

Three things have to stay consistent there, and each fails in its own way: a
reused identifier silently appends to an existing set, a mismatched
`datasetName` fails obscurely, and the id has to be threaded between the calls.

`$products()` collapses all of that into one line, because the search object
already knows both the scenes and the dataset:

```{r}
found$products()
```

Internally it is still the same two calls — `$products()` asks
`$scene_list()` to register, then requests download options against the
resulting id. The identifier is generated (a timestamp plus six random
characters) so it cannot collide, and the dataset name is carried on the object
rather than retyped, so it cannot disagree.

### One registration per search

Repeated calls reuse the same list rather than registering the scenes again:

```{r}
found$products()
found$products()
found$scene_list()$list_id   # all three refer to one registration
```

A narrowed search is a different set of scenes, so it gets its own:

```{r}
found$filter(cloudCover < 10)$scene_list()   # separate list
```

Passing an explicit `list_id` always registers under that name and leaves the
reused one alone, and a list you have deleted with `$remove()` is registered
afresh next time rather than being handed back dead.

Note that nothing deletes these lists automatically — they persist until they
expire, or until you call `$remove()`.

### Working with the list directly

`M2MSceneList` is what you get back when scenes are registered. `$products()`
creates one and uses it without showing you, so most of the time you never
need it. `$scene_list()` gives you the same object to hold onto:

```{r}
found <- sess$dataset("landsat_ot_c2_l2")$search(
  temporal = filter_temporal("2020-07-01", "2020-07-31"),
  max_results = 4
)

scenes <- found$scene_list()
scenes
#> <M2MSceneList>
#>   List id: USGSm2m_20260821150915_vbbem6
#>   Dataset: landsat_ot_c2_l2
#>   Next:    $products()  |  $metadata()  |  $summary()
```

There are three reasons to reach for it.

**Bulk metadata.** `$metadata()` fetches the full metadata for every scene in
the list in a single call. Done scene by scene that would be one request each,
so this is by far the most common reason to want the list:

```{r}
scenes$metadata()
```

**Describing the set.** `$summary()` reports the combined spatial and temporal
extent of everything in the list, and `$scenes()` returns the entity ids it
currently holds:

```{r}
scenes$summary()
scenes$scenes()
```

**Naming a set to come back to.** The identifier is generated for you, but you
can supply your own and reattach to it later in the same session:

```{r}
found$scene_list(list_id = "july_2020_candidates")
sess$scene_list("july_2020_candidates")
```

Bear in mind that lists expire (see below), so this is useful within a working
session rather than across days.

`$remove()` deletes the list from the server when you are finished with it.

### Combining several searches into one set

One search normally means one scene list, so you will not usually think about
what happens when scenes are added to a list that already exists. The exception
is when you want scenes from *more than one* search treated as a single set —
two date ranges, say, or two areas — so that one `$products()` call and one
download order covers all of them.

Give the searches the same `list_id` and they accumulate:

```{r}
ds <- sess$dataset("landsat_ot_c2_l2")

july <- ds$search(temporal = filter_temporal("2020-07-01", "2020-07-05"), max_results = 2)
sept <- ds$search(temporal = filter_temporal("2020-09-01", "2020-09-05"), max_results = 2)

july$scene_list(list_id = "summer_2020")
scenes <- sept$scene_list(list_id = "summer_2020")

scenes$scenes()      # all four scenes
scenes$products()    # products for the whole set
```

Adding **appends** rather than replacing, which is what makes this work, and
re-adding a scene already in the list is a **no-op**, so searches that overlap
will not double-count.

### Lifetime

Scene lists are temporary. They expire server-side after a period of
inactivity, and `$summary()` reports a `listTimeout` for lists that carry one.
Nothing breaks when a list expires — it simply stops being found, and
`$scenes()` returns zero rows rather than raising an error. The same is true
after you delete one with `$remove()`.

This matters when resuming work later: **a scene list will usually be gone by
the next session, but the download order it produced will not.** Reconnect to
orders by label, not to scene lists by id.

### Reattaching to a list

If you reattach to a list by id, the package does not initially know which
dataset it holds — the API only reveals that through the list's summary — so
`$dataset()` resolves it on demand:

```{r}
again <- sess$scene_list("USGSm2m_20260821150915_vbbem6")
again
#>   Dataset: <unresolved>
again$dataset()
#> [1] "landsat_ot_c2_l2"
```

Pass `dataset_name` to `$scene_list()` to skip that lookup, or to choose when a
list spans more than one dataset.

## Choosing products

A scene is not a file. `download-options` reports several *products* per
scene, at two different granularities:

* **Whole products** — a Level-1 Product Bundle (one `.tar` containing the
  entire scene), or a full-resolution browse image. Listed by
  `$scene_products()`.
* **Individual files** — the separate bands and metadata files nested inside a
  bundle. Listed by `$bands()`.

For one Landsat Collection 2 scene that is roughly ten whole products and
nineteen individual files. Browse images have no constituent files, so they
appear only in `$scene_products()`.

Pick a granularity with the matching selector, then narrow with `$filter()`:

```{r}
opts <- found$products()

opts$select_bands(c("B4", "B5"))$filter(bulkAvailable)   # 6 separate .TIFs
opts$select_products("Product Bundle")$filter(available) # 1 .tar per scene
```

`$selected()` always shows what `$request()` will queue. Neither selector
filters on availability, because the API lists superseded products with
`available = FALSE` and silently dropping them would hide that they exist.

## The download queue

`$request()` places an order under a label. A file is ready when it has a URL,
and how it gets one depends on how USGS serves that product:

* **Served by M2M itself** — appears in `$available` with a URL.
* **Proxied** — appears in `$requested`, but *also* with a URL. These are
  hosted elsewhere (`landsatlook.usgs.gov`, for instance) on a signed URL, so
  M2M hands you the link and steps out of the way. They are immediately
  downloadable.
* **Still staging** — appears in `$requested` with no URL yet.

Because the API's two buckets do not line up with ready-versus-pending, use
`$ready()` and `$pending()` rather than reading `$available` and `$requested`
directly. `$is_ready()` is true when nothing is left staging, and `$refresh()`
re-polls:

```{r}
queue <- opts$select_bands(c("B4", "B5"))$request(label = "ndvi_july_2020")

while (!queue$is_ready()) {
  Sys.sleep(30)
  queue$refresh()
}

queue$retrieve("data/landsat")
```

`$refresh()` accumulates rather than replaces, because the API drops entries
once it has handed them over — a plain replacement would lose URLs collected on
an earlier poll.

Some orders are staged rather than queued, and need `$prepare()` to start
processing. Along with `$cancel()`, it is one of the two methods here that
change server-side state.

## Coming back later

Orders outlive both your R session and the scene list that produced them, so a
large download does not have to finish in one sitting:

```{r}
sess <- m2m_session()

sess$download_labels()      # which orders exist
queue <- sess$download_queue("ndvi_july_2020")
queue$refresh()
queue$retrieve("data/landsat")
```

`sess$downloads()` lists every individual download across all labels, and
`queue$summary()` breaks one order down by dataset.

## What lives where

| Thing | Identified by | Survives your session? | Created by |
|---|---|---|---|
| Session | API token | No — expires when idle | `m2m_session()` |
| Scene list | `listId` | Usually not — expires when idle | `$scene_list()`, or implicitly by `$products()` |
| Download order | `label` | Yes | `$request()` |
| Downloaded files | path on disk | Yes | `$retrieve()` |

The practical consequence: give orders labels you will recognise later, and do
not expect a scene list to still be there tomorrow.
